Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models

### 2015-02-07 processes-69103_review_response/processes-69103.tex
Changes to tex
Changes

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa},
and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
\caption{Reduced order coagulation model training simulations.
Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) without the protein C pathway as a function of prothrombin. 
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
(A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of protein C pathway.
The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
As depicted in (A,B,C) the prothrombin levels were at 150\%, 100\% and 50\% of their physiological concentration in the absence of protein C pathway. 
All factors and control proteins in these experiments were at their physiological concentration unless otherwise denoted.}\label{fig-diff_fII_NO_APC} 
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
\caption{Reduced order coagulation model training simulations.
Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with the protein C pathway as a function of prothrombin.
Only APC pathway parameters were allowed to vary in these simulations keeping the parameters estimated without protein C pathways constant. 
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
(A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
As depicted in (A,B,C) the prothrombin levels were at 150\%, 100\% and 50\% of their physiological concentration in the presence of protein C pathway. 
All factors and control proteins in these experiments were at their physiological concentration unless otherwise denoted.}\label{fig-diff_fII_APC}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
The reduced order coagulation model parameter estimates were tested against data not used during model training.
Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}.
Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
As depicted in (A,B,C,D) the prothrombin and ATIII levels were at (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of their physiological concentrations in the absence of protein C pathway. All factors and control proteins were at their physiological concentration unless otherswise denoted.}\label{fig-fII_ATIII}
\end{figure}

@@

	f_{i}\left(\mathcal{Z}_{j}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
$\kappa_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).

@@

Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
All plots were made using the Matplotlib module of Python \citep{Matplotlib}.
%The optimization and plotting code can be downloaded from http://www.varnerlab.org.

@@

@@

\end{document}

>>>>

Previous coagulation models have typically been formulated as systems of nonlinear ordinary differential equations, 
using mass action or more complex kinetics, to describe the rates of biochemical conversions \citep{Khanin:1989aa,Willems:1991aa,Baldwin:1994aa,Leipold:1995aa,Kuharsky:2001aa}.
Mechanistic ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys} 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, Hockin \emph{et al.} \citep{Hockin:2002aa}, and later Butenas et al., \citep{Butenas:2004aa}
who developed and then subsequently refined highly mechanistic coagulation models. Recently Papadopoulos and co-workers used a phenomenological  mathematical model for thrombin generation \cite{Atkin:2014}. Using a set of four ordinary differential equations they were able to derive an equation for temporal evolution of thrombin generation. However the model focuses more on thrombin generation rather than dynamics of coagulation. It lacks any detail regarding regulatory activity and is unable to capture the roles of primary inhibitors like ATIII or protein C as the hybrid approach captures.The model also requires that the parameters are adjusted or fine tuned for each data set. This makes the model unfeasible in cases where we need to understand the dynamics of coagulation without complex modeling approaches or when there is incomplete mechanistic knowledge. Other laboratories have also expanded upon Hockin \emph{et al.}, for example by exploring the intrinsic pathway, the role of stochastic fluctuations in coagulation \citep{Lo:2005aa}, and the dynamics of thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside clots \citep{Voronov:2013aa}.
However, these previous studies were largely based upon extensive mechanistic knowledge. 
This is possible because blood, while enormously complex, can be systematically interrogated.
Other systems, such as intracellular signaling networks, are much more difficult to experimentally interrogate.  
Towards this unmet need, we formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which a complete mechanistic understanding is missing. 
We tested this approach by modeling the human coagulation cascade. 
The hybrid model consisted of only five differential equations augmented with several logical rules.
Thus, the model was more than an order of magnitude smaller than comparable purely ODE models in the literature.
We estimated the model parameters from \textit{in~vitro} extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then compared the model predictions with thrombin data sets, for both normal and hemophilic coagulation, that were not used for model training.
Once validated, we performed flux and sensitivity analysis on the model to estimate which parameters were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, 
suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-3-DIFF_FII_NO_APC.pdf}
\caption{Reduced order coagulation model training simulations.
Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) without the protein C pathway as a function of prothrombin. 
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data. 
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
(A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the absence of protein C pathway.
The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}\citep{Veer:1997}.
Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
As depicted in (A,B,C) the prothrombin levels were at 150\%, 100\% and 50\% of their physiological concentration in the absence of protein C pathway. 
All factors and control proteins in these experiments were at their physiological concentration unless otherwise denoted.}\label{fig-diff_fII_NO_APC} 
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-4-DIFF_FII_APC.pdf}
\caption{Reduced order coagulation model training simulations.
Reduced order coagulation model parameters were estimated using particle swarm optimization (PSO) with the protein C pathway as a function of prothrombin.
Only APC pathway parameters were allowed to vary in these simulations keeping the parameters estimated without protein C pathways constant. 
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation).
(A,B,C) depict training data and results for 150\%, 100\% and 50\% of physiological prothrombin levels in the presence of the protein C pathway. 
The experimental training data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}\citep{Veer:1997}.
Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
As depicted in (A,B,C) the prothrombin levels were at 150\%, 100\% and 50\% of their physiological concentration in the presence of protein C pathway. 
All factors and control proteins in these experiments were at their physiological concentration unless otherwise denoted.}\label{fig-diff_fII_APC}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-FII_ATIII.pdf}
\caption{Reduced order coagulation model predictions versus experimental data for normal coagulation. 
The reduced order coagulation model parameter estimates were tested against data not used during model training.
Simulations of different levels of prothrombin and ATIII were compared with experimental data in the absence of the protein C pathway. 
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles, points denote experimental data.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
(A,B,C,D) prediction results for (FII,ATIII): (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of physiological prothrombin and ATIII levels in the absence of the protein C pathway. The experimental validation data was reproduced from the study of Butenas \textit{et al.} \citep{Butenas:1999aa}\citep{Veer:1997}.
Thrombin generation in these experiments was initiated using 5 pmol/L FVIIa-TF in the presence of 200 $\mu$mol/L of phospholipid vesicles (PCPS). 
As depicted in (A,B,C,D) the prothrombin and ATIII levels were at (50\%,150\%), (100\%, 100\%), (125\%, 75\%) and (150\%, 50\%) of their physiological concentrations in the absence of protein C pathway. All factors and control proteins were at their physiological concentration unless otherswise denoted.}\label{fig-fII_ATIII}
\end{figure}

@@

	f_{i}\left(\mathcal{Z}_{j},k_{ij}\right) = \frac{k_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + k_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $k_{ij}$ and $\eta$ are control parameters. 
$k_{ij}$ was the species gain parameter, while $\eta$ was a cooperativity parameter (similar to a Hill coefficient).

@@

Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language. 
All plots were made using the Matplotlib module of Python \citep{Matplotlib}.\newline
%The optimization and plotting code can be downloaded from http://www.varnerlab.org.
%Experimental data,simulation code,parameter ensembles and initial conditions for each experimental simulation can be viewed and downloaded from https://github.com/jeffreyvarner/HybridCoagulationModel\_v1.

@@

\bibitem{Atkin:2014}
Papadopoulos, Konstantinos and Gavaises, Manolis and Atkin,Chris (2014) A simplified mathematical model for thrombin generation.
\newblock Medical Engineering \& Physics 36: 196 - 204.

\bibitem{Veer:1997}
Cornelis van't Veer, Neal J. Golden, Michael Kalafatis and Kenneth G. Mann (1997) Inhibitory Mechanism of the Protein C Pathway on Tissue Factor-induced Thrombin Generation.
\newblock The Journal of Biological Chemistry 272: 7963-7994.

@@

\bibliography{References_v1.bib}
\end{document}
